---
title: "Clean Up"
author: "Ben Elam"
date: "March 1, 2016"
output: html_document
---
First of all, we're going to be really, really unhappy until we break this up into different data sets.
Remember that problem with missing values becoming strings and strings becoming factors?
It's happening here!

We have NA values, missing values, and even a value of "Na". This mixed case will be treated as a string, not as an NA value.

```{r}
set_sp_grp = function(species) {
  if (species %in% c('Q.stellata', 'Q.alba')) # Post and white
    return('WhiteOak')
  else if (species %in% c('Q.shumardii', 'Q.pagoda')) # Shumard and cherrybark
    return('RedOak')
  else if (species %in% c('C.ovata', 'C.glabra')) # Shagbark and pignut hickories 
    return('Carya')
  else return(NA)
}

set_sp_grp = Vectorize(set_sp_grp, vectorize.args = 'species')

setwd('~/work/nmsds/guides/showandtell/atf')
atf = read.csv('ATFMangled.csv',
               stringsAsFactors = FALSE,
               na.strings = c('NA', 'Na', '')) # Skip this step and you're gonna have a bad time.

atf$SpGrp = set_sp_grp(atf$Species)
```

Note that the procedure is the same for all of the ions.
Just change the column names as appropriate.
We'll restrict our investigation to the chloride data.

```{r}
#clnames = c('Date', 'PgTF', 'Species', 'Cl', 'ClER')
clnames = c('Date', 'Species', 'Cl', 'ClER', 'SpGrp')
cl = atf[atf$Species != 'precipitation', clnames]
names(cl)[names(cl) == 'ClER'] = 'ER'
```

Now, we would like to do the following:

- Split our chloride data set, `cl`, into a list of data frames by date.
- Remove data frames with NA values.
- Merge the good data frames back together.

To accomplish this, we'll first write a function that determines
whether or not a data frame contains an NA entry.

```{r}
no_NA = function(df) {return(!as.logical(sum(colSums(is.na(df)))))}

cl_dates = split(cl, cl$Date)
cl_dates_clean = cl_dates[sapply(cl_dates, no_NA, simplify = 'vector')]
# Because `cl_clean = rbind(cl_dates_clean, make.row.names=FALSE)` is not what we want:
cl_clean = do.call("rbind", as.list(c(cl_dates_clean, make.row.names=FALSE)))
```

We can aggregate our species group data easily enough:

```{r}
sp_grp_er = aggregate(ER ~ Date + SpGrp, cl_clean, mean)
names(sp_grp_er)[names(sp_grp_er) == 'ER'] = 'ERSpGrp'
sp_grp_er = sp_grp_er[order(sp_grp_er$Date),]
sp_grp_er
```

That's great, but how do we get this data into our `cl_clean` data frame?

```{r}
merged = merge(cl_clean, sp_grp_er, by = c('Date', 'SpGrp'))
merged = merged[, c('Date', 'Species', 'Cl', 'ER', 'SpGrp', 'ERSpGrp')]
head(merged)

#write.csv(merged, 'data/Chloride.csv', row.names = FALSE)
```