---
title: "R Data Structures"
author: "Ben Elam"
output: html_document
---
R comes with a number of functions for data analysis and modeling
that work out of the box.
Many more are available through its primary package library, CRAN.
This workshop is not about those functions.

On the one hand, some of these tools can nearly complete your thesis for you.
On the other, they're not worth a lot if your data is in the wrong format.
This workshop is about developing an understanding of how R represents data
in order to efficiently format our data inputs and extract our results.

We'll step through a brief example which will open up some nice questions,
which we'll then try to provide some rather general answers to.

## A Motivating Example
Let's do something simple. Let's make a copy of a dataset, `trees`, that ships with R.
(For a list of all such data sets, enter `data()` in the REPL.)

```{r}
head(trees)
```

Suppose we're interested in the distribution of tree heights.

```{r, echo=FALSE}
ggplot2::qplot(trees$Height, geom="histogram", bins=5) +
  ggplot2::ggtitle("Trees") +
  ggplot2::xlab("Height") +
  ggplot2::ylab("Count")
```

Now, we'd like to get the cumulative frequency distribution of the `Height` column.
To do so, we'll want our copy of `trees` to be ordered by `Height.
We'll also reorder the columns themselves so that `Height` is to the right of the others.

```{r}
df = trees[order(trees$Height), # Reorder rows by Height
           c('Girth', 'Volume', 'Height')] # Reorder columns as described
```

This is a fairly simple line of code, but there's still a lot going on!

- We subset the data frame `trees` by both row and column
using the `[,]` syntax.
Note that we aren't losing any elements by doing so.
- We subset a single vector of `trees` using `trees$Height`.
- We use the `order()` function.
It returns a vector of the indices of `trees$Height` after they've been sorted,
**not** the elements themselves.
This vector defines the row numbers of the new data frame.
- We provide a vector of column names in the order we prefer.

So, let's glance at what we got back:
```{r}
head(df)
```

Things look good with our data set now, so we're ready to move on.

### Derived Data
Since the heights are ordered, we can compute their cumulative sums.
R has a nice built-in function for this - `cumsum()` -
that does just what we want.

```{r}
total_height = sum(df$Height)
df$Height_Cum = cumsum(df$Height)
df$Height_Prop = df$Height_Cum / total_height
head(df)
```

Here, we use a couple of functions - `sum()` and `cumsum()`.
We also assign two new columns to our data frame.

## Data Structures In Brief
Wikipedia defines a
[data structure](https://en.wikipedia.org/wiki/Data_structure) as follows:

> In computer science, a data structure is a particular way of organizing data
in a computer so that it can be used efficiently.

In R, we're mostly concerned with data structures known as
vectors, lists, and data frames.
For particularly computational work,
an additional structure - an array - is often best.
However, the flexibility of data frames tends to better suit most analyses.

To understand data frames, it's important to first understand
the data structures underpinning them.

### Vectors
R represents sequential collections with **vectors**.
There are two kinds of vectors: **atomic vectors** and **recursive vectors**.
Since this terminology is a bit unwieldy,
atomic vectors are simply called vectors,
and recursive vectors are called **lists**.
Similarly, when people talk about vectors,
they are usually referring to atomic vectors.

While each has a few additional idiosyncrasies,
there are three differences between vectors and lists that are critical to observe.

First, lists may contain a mixture of data types,
but all elements of a vector must be of one data type.
If a vector is created from objects of more than one type,
then R will coerce all elements to the most flexible type present.
In terms of flexibility, string > numeric > integer > bool.
For instance:

```{r}
v = c(1, 2, 'three')
typeof(v)
print(v)
```

Of course, the above example isn't so likely to occur.
A recording error during data collection, however, might be a problem.
Suppose that the following data was loaded with `df = read.table('xy.txt')`:
```
x y
1 2.53
2 8,81
3 7.20
```

The usage of a comma as a decimal delimiter is not uncommon outside of the USA,
and perhaps someone used to both systems accidentally made the above error when
recording data from their lab notebook.
How does R respond?

- Try to create a data frame.
- Realize the elements of `y` are not homogeneous.
- "8,81" is not a numeric type, so load it as a string.
- Make every element of `y` a string.
- Load string vector as a Factor (unless `stringsAsFactors=F` was used.)

Now, we have a data frame with a categorical variable where we expected a
numerical variable! This is probably not what we wanted.

Second, lists do not support **vectorized operations**.
This is easiest to communicate by example.
This works:

```{r, eval=FALSE}
v = c(1, 2, 3)
v * 2 # Returns c(2, 4, 6)
v + c(3, 2, 1) # Returns c(4, 4, 4)
```

However, this would produce an error:

```{r, eval=FALSE}
l = list(1, 2, 3)
l * 2 # Error in l * 2 : non-numeric argument to binary operator
```

Finally, vectors are flat structures.
```{r, eval=FALSE}
c(c(1, 2), c(3, 4)) # Equivalent to c(1, 2, 3, 4)
c(c(1, 2), list(3, 4)) # Enforces homogeneity, then flatness. Equivalent to list(1, 2, 3, 4)
c(list(1, 2), list(3,4)) # Same as above
```

```{r}
ll = list(c(1, 2), c(3, 4)) # Exactly what you asked for - a list of two vectors.
str(ll)
```

So, how do we access elements of vectors and lists?

### Named Structures & Structure Preservation
Both vectors and lists support naming of their elements.
Data frames, which we'll see shortly, actually enforce naming by default.

```{r}
v = c('a' = 1, 'b' = 2)
```

Naming can make accessing our data much nicer.
Subsetting an object by index or by name using square brackets
will **preserve** the structure of the original object.
In the cases below, both return the result as a named vector.

```{r, eval=FALSE}
v[1] # Returns c('a' = 1)
v['a'] # Also returns c('a' = 1)
v[c('b', 'a')] # Returns c('b' = 2, 'a' = 1)
```

If we instead use two pairs of brackets, R will choose the data structure
that best **simplifies** the organization of the data without
losing the data itself.
The subsetting operations below only return a singleton vectors.

```{r, eval=FALSE}
v[[1]] # Returns 1
v[['a']] # Returns 1
v[[c('b', 'a')]] # Error ... : attempt to select more than one element
```

The preceding subsetting commands would also work if `v` was a list.
However, lists have an additional operator -
`$` - for accessing named elements.
This symbol is simply shorthand for using doubled square brackets.
```{r, eval=FALSE}
l = list('x' = c(1, 2), 'y' = c(3, 4))
l['x'] # Returns list('x' = c(1, 2))
l[['x']] # Returns c(1, 2)
l[['y']][2] # Returns 4
l$y # Returns c(3, 4)
```

### Arrays
An **array** is a vector with an additional attribute - a dimension - named `dim`.
This makes it possible to circumvent the flatness property of atomic vectors,
which have `NULL` dimension.
An array is effectively a vector that can contain other vectors up to some arbitrary dimension,
but it's more accurate to say that an array contains other arrays.

A normal atomic vector `v` has dimension `NULL`.

```{r}
dim(c(1,2,3))
```

A vector with a `dim` attribute is a one-dimensional (1D) array, and a matrix is a 2D array.
The following will all produce the same 3x3 matrix:

```{r}
M = 1:9 * 2
dim(M) = c(3, 3)

M = array(1:9 * 2, c(3, 3)) # Use the built-in constructor for general arrays

M = matrix(1:9 * 2, ncol=3, nrow=3) # Built-in constructor for matrices
```

There is certainly more to say about arrays.
However, it's sufficient for now to remark that they support vectorized
operations and enforce homogeneous typing, just like a 1D vector.
For standard data analysis that doesn't require much in the way
of linear algebra or physics,
a more flexible data structure exists in R.

### Data Frames
A **data frame** is a list of vectors.
Cross-sections of vectors can be accessed by row names,
so a data frame functions like a table.
Naming of columns and rows isn't exactly required,
but it's recommended.

```{r, eval=FALSE}
df = read.table('data.txt')
```

```{r}
df = data.frame(numbers =  c(1:4), characters = sample(letters, 4))
df[1] # Get first column
df[c(1, 2), 'characters'] # Get 1st and 2nd row of characters column
```

In the last command above,
we used a single set of square brackets to tell R
to preserve the structure of the data frame.
Then, we divided two arguments with a comma.
The first was a vector of indices specifying rows to select data from,
and the second was the name of a column to select from.
Of course, multiple columns can be selected by providing them as a vector,
as demonstrated in the motivating example.

With this in mind, it follows that each of the following will extract a column:

```{r, eval=FALSE}
df[, 'numbers'] # Returns c(1, 2, 3, 4)
df[['numbers']] # Returns c(1, 2, 3, 4)
df$numbers # Returns c(1, 2, 3, 4)
```