---
title: "NMDS Workshop1"
author: "Ben Elam"
date: "February 1, 2016"
output: html_document
---

## This Workshop
Goals:

- Provide new users with an idea of how R might be of use to them.
- Make sure that all workshop attendees (new and prior) have a baseline background that can be built upon in future workshops.

Visit [github.com/nmsdataskills/r_basics](https://github.com/nmsdataskills/) to find the latest version of this guide.
You can download the RMarkdown (.Rmd) file to play with in RStudio,
read or download the HTML version of the guide,
or just download a simplified R (.r) script with the code.
You'll also want the data (TODO) to follow along throughout the workshop.

This workshop is not:

- A general primer on programming
- An introduction to statistics or even to statistical programming

See Resources (at bottom) for links to various things we recommend reading or watching.

## North Mississippi Data Skills
We provide accessible training in professional, open-source data workflows and best practices to researchers from various backgrounds.

Our workshops are designed for students with research projects in mind.
We assume that your program of study involves some statistical preparation relevant to your research.
From experience, we also assume that your instructor had to cover a lot statistics,
and only had so much time to cover professional tools.
That's where we hope to be of assistance!

Where to find us:

- Hang out with us at [nmsdataskills.slack.com](http://nmsdataskills.slack.com). Currently, anyone with an msstate.edu email address can join the team.
- Access our content at [github.com/nmsdataskills](http://github.com/nmsdataskills). We're happy to have contributions.

## Thanks!
We'd like to thank the MSU Mitchell Memorial Library and the MSU Instructional Media Center
for providing us with this meeting space.

## Loading Data
### The `setwd` Command
```{r}
setwd("~/work/nmsds/guides/R_intro/")
```

### Filepaths
On any platform, R adheres to the Unix format of delimiting directories in a pathname by forward slashes - like this `/`, not like this `\`.

This might confuse Windows users, whose filepaths usually look something like `C:\Users\ben\work\nmsds\guides\R_intro`. To use this filepath in R, you would enter
```{r, eval=FALSE}
setwd("C:/Users/ben/work/nmsds/guides/R_intro/")
```

### Headers
no header
header
multi-line header

## Data At A Glance
Get the number of rows in a dataframe: `n = nrow(df)`


### Printing A Dataframe
### The `summary` Command
### Visualization With The `plot` Command
```{r, echo=FALSE}
plot(cars)
```

## Basic Data Manipulation and Cleaning
### Understanding Boolean Expressions

### Subset
Syntax

Example selections

Summarize again

Plot again

## Generating Mock Data With Random Numbers
Don't have your data yet? Not a problem!

R has some nice commands for generating random data. For example:
```{r, echo=TRUE}
set.seed(1) # Get rid of this to get a new data set each time.
runif(4, min=-1, max=1) # Generate 4 random floating point numbers between -1 and 1.
rnorm(3) # Generate 3 floating point numbers from the N(0,1) distribution
rnorm(5, mean=70, sd=10) # Generate 5 floating point numbers from the N(70,10) distribution
sample(1:10, 4) # Generate 4 integers between 1 and 10 without replacement
```

So, why would we do this?

- You can write code prior to receiving your data, which is important if your analysis is time-sensitive or if your data set is updated over time.
- You think about the possible values your observations could take on. What does it mean for an observation to be zero valued? Negative?
- By designing in advance, you're restraining yourself from designing your analysis to suit your data. Generally, we'd like to design an analysis in advance, then arrange for data to be collected in a manner that ensures the assumptions of our analysis will be met.

So, the goal is not to make inferences about the data,
but to make sure that our code can handle the format of the data we expect.


### Creating The `newcars` Data Set
This is how the `newcars` data set was generated from the "cars" data set
that is distributed with R.

```{r}
set.seed(42)
newcars = cars
newcars$label = sample(1:4, nrow(newcars), replace=TRUE)

write.csv(newcars, file="newcars.csv")
# Save tbl

```

Notice that it wasn't necessary to specify a full file path when writing my data to each file.
The directory specified above with `setwd` was used.
If the files need to be stored outside of the current working directory,
`write.csv` and `write.table` both work fine with full pathnames:
```{r, eval=FALSE}
write.csv(newcars, file="~work/nmsds/guides/fake-directory/newcars.csv")
```

## Resources
The first place you should look for anything R-related is the
[Awesome R](https://github.com/qinwf/awesome-R) collection.
For similar lists, see [Awesome Awesomeness](https://github.com/bayandin/awesome-awesomeness).

### Getting Set Up
You can learn about installing R at [CRAN](https://cran.r-project.org/) - the Comprehensive R Archive Network. This presentation was made using
[RStudio Desktop](https://www.rstudio.com/products/rstudio-desktop/)
and RMarkdown.

RStudio makes [great reference cards](https://www.rstudio.com/resources/cheatsheets/)
for their products. You might be interested in:

- [Using RStudio](http://www.rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf)
- [RMarkdown Reference](http://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [RMarkdown Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [Data Wrangling with dplyer and tidyr](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
- [Visualization with ggplot2](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)

### Learning R
CodeSchool has some great introductory lessons to things like Git and web development. They're picking up on some other topics, too, [like R](http://tryr.codeschool.com/)!

[Datacamp](https://www.datacamp.com/) is supposed to be pretty good. I can't say I've tried it.

For a quick overview, you might like the Learn X in Y Minutes project, which has [an entry on R](https://learnxinyminutes.com/docs/r/).

CRAN provides a [reference card](https://cran.r-project.org/doc/contrib/Short-refcard.pdf)
for the R language.

Of course, you could always just Google "R MOOC"...